Computing the bicharacter of huge subspaces of polynomials

### 2019-05-13 article/article.tex
Complete section on repr. of Sn

\DeclareMathOperator{\Sn}{\mathbb{S}_n}
\DeclareMathOperator{\GLr}{GL_r}
\DeclareMathOperator{\harmonics}{\mathcal{H}}

@@

	Consider the polynomial ring $P=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables: \pauline{small $x$ for variables, capital $X$ for sets of variables? } \pauline{introduce inert variables now or later?}
	$$(X_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $X_i = (x_{i1}, x_{i2}, \dots, x_{in})$ the i-th set of variables.

@@

	It is known (\cite{Haiman2002} and \cite{Bergeron2009}) that the harmonics $\harmonics_{n,1}$ are generated by the Vandermonde determinant $V_n$ and closed by partial derivatives. Moreover, the diagonal harmonics $\harmonics_{n,r}$ are generated by $\harmonics_{n,1}$ and closed by the polarization operators.

@@

	In their work, Garsia and Haiman also introduced a generalized version of the Vandermonde determinant and a generalization of the diagonal harmonics for two sets of variables called the Gasia-Haiman modules. \\

@@

	We first need to compute in a quotient by an ideal.
	A standard approach would be to compute a Gröbner basis of this ideal.
	We haven't actually tried it, but given the number of variables and dimension of the quotient, we expect this calculation to be intractable.

	So we took the classical route of realizing the coinvariant ring as the space $\harmonics_{n,r}$ of diagonal harmonic polynomials.
	This is the space of those polynomials that satisfy the equations given by the multisymmetric functions seen as differential operators.

@@

	\begin{theorem}
		Diagonal harmonic polynomials are obtained by polarization of harmonic polynomials in the first set of variables.
	\end{theorem}

	The space of harmonic polynomials $\harmonics_n$ itself can be constructed, for example, as the subspace spanned by the Vandermonde determinant under differentiation.

	In practice, we implemented a general purpose utility in Sage \texttt{Space(vectors, operators)} that computes the smallest subspace of a vector space which contains the given vectors and is stable under the action of the given linear operators.
	\nicolas{put an example here}

@@

@@

	To get a basis of the full isotypic component $\harmonics^\lambda$, we can, for example, use a generalization of this construction (voir \cite{Ariki1997}): a family of polynomials $(G_{P,Q})_{P,Q}$ indexed by a pairs of standard tableaux of shape $\lambda$:
	- $Q$ specifies the copy of $V_\lambda$
	- $P$ specifies the location within that copy of $V_\lambda$

	Remark: For a fixed tableau $P$, denote by $\harmonics_k^P$ the space spanned by the higher Specht polynomials $(G_{P,Q})_Q$ under polarization. It's exactly the image of $\harmonics$ under the Young idempotent associated to $P$. For a given shape $\lambda$, the $\harmonics_k^P$ are all conjugate under $\Sn$, and their direct sum is the isotypic component $\harmonics_k^\lambda$.

	Conclusion: It is sufficient to compute $\harmonics_k^P$ for one tableau of each shape $\lambda$. This reduces the complexity by the square of the number of standard tableaux of shape $\lambda$

@@

	The polarization operators skew commute. \pauline{add commutator computation} Therefore it would be in principle sufficient to apply them in some "decreasing order". This sounds promising given that, at this stage about 90\% of the polynomials that are produced are redundant and reduced to zero by the linear algebra.

@@

	\section{Applications}

>>>>

\DeclareMathOperator{\Sn}{\mathbb{S}_n}
\DeclareMathOperator{\GLr}{GL_r}
\DeclareMathOperator{\harmonics}{\mathcal{H}}
\DeclareMathOperator{\sgn}{sgn}

@@

	\subsection{Harmonic and diagonal harmonic polynomials}

	Consider the polynomial ring $P=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables: \pauline{small $x$ for variables, capital $X$ for sets of variables? }
	$$(X_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $X_i = (x_{i1}, x_{i2}, \dots, x_{in})$ the i-th set of variables.

@@

	It is known (\cite{Haiman2002} and \cite{Bergeron2009}) that the harmonics $\harmonics_{n,1}$ are generated by the Vandermonde determinant $V_n$ and closed by partial derivatives. 
	%Moreover, the diagonal harmonics $\harmonics_{n,r}$ are generated by $\harmonics_{n,1}$ and closed by the polarization operators.

@@

	\subsection{Generalizations}

	In their article \cite{GarsiaHaiman1993}, Garsia and Haiman also introduced a generalized version of the Vandermonde determinant and a generalization of the diagonal harmonics for two sets of variables called the Gasia-Haiman modules. \\

@@

	We first need to compute in a quotient by an ideal, $P / I_n$ where $I_n$ is the ideal of diagonal symmetric polynomials on $r$ sets of $n$ variables. 
	A standard approach would be to compute a Gröbner basis of this ideal.
	We haven't actually tried it, but given the number of variables and dimension of the quotient, we expect this calculation to be intractable.

	So we took the classical route of realizing the coinvariant ring as the space $\harmonics_{n,r}$ of diagonal harmonic polynomials.
	This is the space of those polynomials that satisfy the equations given by the multisymmetric \pauline{multi symmetric vs diagonal symmetric ?} functions seen as differential operators.

@@

	\begin{theorem} \label{DiagHarm}
		Diagonal harmonic polynomials are obtained by polarization of harmonic polynomials in the first set of variables.
	\end{theorem}

	Therorem \ref{DiagHarm} has been proven for two sets of variables. It is not proven yet for more than two sets of variables but it has been verified by calculation for a large number of sets of variables so we assume it to be true since we will never reach that number of sets in our case.

	The space of harmonic polynomials $\harmonics_n$ itself can be constructed, for example, as the subspace spanned by the Vandermonde determinant under differentiation. \\

	In practice, we implemented a general purpose utility in Sage, \texttt{Subspace(vectors, operators)} that computes the smallest subspace of a vector space which contains the given vectors and is stable under the action of the given linear operators.
	\nicolas{put an example here}

@@

	In practice, first we compute $\harmonics_{n}$ and decompose it into isotypic components using the Young symmetrizer. Then we use the polarization operators to obtain the isotypic components of $\harmonics_{n,r}$ from those of $\harmonics_{n}$. \\

	To define the Young symmetrizer, we first need to define two elements of the symmetric group algebra $\CC \Sn$. Let $P_{\lambda}$ be the subgroup of $\Sn$ of all row preserving permutations and $Q_{\lambda}$ the subgroup of the columns preserving permutations. Then we have $a_{\lambda} = \sum_{\sigma \in P_{\lambda}} \sigma$ and  $b_{\lambda} = \sum_{\sigma \in P_{\lambda}} \sgn(\sigma) \sigma$. Finally, the Young symmetrizer is the element $$c_{\lambda} = a_{\lambda} b_{\lambda}.$$

	\begin{theorem} \cite{FultonHarris1991}
		The Young symmetrizer $c_{\lambda}$ is idempotent up to a scalar and the image of $c_{\lambda}$ by right multiplication on $\CC \Sn$ is an irreducible representation $V_{\lambda}$ of $\Sn$.
		Thus any irreducible representation can be obtain this way for a unique partition. 
	\end{theorem}

	This theorem guarantees that $c_{\lambda}$ is an isotypic projector and that applying it on the elements of $\harmonics_{n}$ gives representatives of the isotypic components of $\Sn$. \pauline{Do we need to prove explicitly commutation with symmetric operators?}

	As mentioned above, since the action of $\GLr$ and $\Sn$ commutes, at this stage we only have to apply the polarization operators on each isotypic component independently to be able to compute the bicharacter. \\

@@

	To get a basis of the full isotypic component $\harmonics^\lambda$, we can, for example, use a generalization of this construction (see \cite{Ariki1997}): a family of polynomials $(G_{P,Q})_{P,Q}$ indexed by a pairs $(P,Q)$ of standard tableaux of shape $\lambda$ where $Q$ specifies the copy of $V_\lambda$ and $P$ specifies the location within that copy of $V_\lambda$.

	Remark: For a fixed tableau $P$, denote by $\harmonics_k^P$ the space spanned by the higher Specht polynomials $(G_{P,Q})_Q$ under polarization. It's exactly the image of $\harmonics_n$ under the Young idempotent associated to $P$. For a given shape $\lambda$, the $\harmonics_k^P$ are all conjugate under $\Sn$, and their direct sum is the isotypic component $\harmonics_k^\lambda$.

	To conclude, it is sufficient to compute $\harmonics_k^P$ for one tableau of each shape $\lambda$. This reduces the complexity by the square of the number of standard tableaux of shape $\lambda$

@@

	The polarization operators skew commute. Therefore it would be in principle sufficient to apply them in some "decreasing order". This sounds promising given that, at this stage about 90\% of the polynomials that are produced are redundant and reduced to zero by the linear algebra.

@@

	\section{Applications}
	Ask Francois ...
